\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-02-15 document/XSGRB_sample.tex
Last updates before submission

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

\usepackage{natbib}
\usepackage{caption}

@@

\usepackage{float}

@@

\title{XS-GRB I - The X-shooter GRB afterglow legacy sample \thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Program IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811, 096.A-0079,
		097.A-0036, and 098.A-0055.}}

\titlerunning{The X-shooter sample of GRBs}

@@

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. This totals spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneous sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results, that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

\keywords{Gamma-ray burst: general --- galaxies: high-redshift --- ISM: molecules --- dust, extinction }

@@

$\gamma$-ray bursts (GRBs) are bright glimpses of electromagnetic radiation that
pierce through the universe, all the way from the beginning of time. They
provide constraints on a very wide range of topics in astrophysics. Examples
range from small-scale phenomena relating to magnetars, properties of highly
relativistic jets, hyper/supernova explosions, the interstellar medium, dust
extinction curves, starburst galaxies, chemical and molecular abundances, escape
of ionizing radiation, the ionization state of the intergalactic medium,
intervening absoption systems to standard candles in cosmology
\citep[e.g.,][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

The \textit{Niel Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the two on-board
instruments, the Burst Alert Telescope (BAT) \citep{Barthelmy2005} and the X-Ray
Telescope (XRT) \citep{Burrows2005}, \swift~is an ideal telescope for GRB
hunting. A crucial aspect of the success of the {\it Swift} mission has been the
extensive ground-based follow-up observations of the afterglows and of the host
galaxies of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the
{\it Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB180817A/AT2017gfo
\citep{LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

In the beginning of the {\it Swift} era most of the follow-up afterglow
spectroscopy was secured using low-resolution spectrographs \citep[typically
$R=\lambda/\Delta\lambda$$<$1000, e.g.,][]{Fynbo2009}. Spectroscopy is powerful
as it allows us to secure information even for very faint targets (24.5
\citep{Kruhler2012}). This allows the measurements of a number of important
parameters such as redshifts, spectral slopes, and extinction. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra have been
secured, and for these events much more information about conditions inside the
host galaxies were extracted \citep[e.g.,][]{Fiore2005, Thone2007,
	Prochaska2007, Vreeswijk2007, DElia2009, Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} was the first of the second
generation instruments at the ESO Very Large Telescope (VLT). It was designed
very much with transient source follow-up in mind as the fading luminosities of
such sources makes it urgent to secure as extensive coverage as possible in the
shortest possible time. At the same time, the resolution was designed to be in
the range 4000--9000 in order to be able to get a large useful spectral range
between the many sky-background emission lines in the red and near-IR spectral
ranges. The large spectral coverage allows for the spectroscopic observations of
the highest redshift GRBs.

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all data resulting from the survey, publicly available in reduced form (see Sect
\ref{products}).

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also here assess the
completeness of the realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}. We use the $\Lambda$CDM cosmology parameters
provided by \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs without hampering the follow-up effort. Many natural follow-up
restrictions exist already, being it weather conditions, pointing restrictions
of the telescope or unconstrained burst localizations as reported by the alerting
telescope. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitate an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

In defining the selection criteria, we simultaneously aim to minimize any biases
against intrinsic astrophysical conditions while at the same time maximizing the
likelihood of successful observations, hence allowing us to obtain a higher
degree of completeness. By restricting the selection criteria to conditions
local to the Milky Way and therefore independent of intrinsic GRB properties,
the aim is that the sample collected represents the underlying distribution of
GRBs in a fair way. The selection criteria used here are based on previous,
similar studies \citep{Jakobsson2006b, Fynbo2009, Hjorth2012}. We characterize
the sample completeness in Sect. \ref{completeness}.

The selection criteria that define a GRB as part of our \textit{statistical}
	sample are:

@@

	\item GRB triggered by BAT onboard the \swift~satellite

@@

	\item The target must be visible for at least 60 minutes, 30 degrees above the
	horizon, with the Sun below $-12$ degrees\footnote{Note that in the P84
		proposal the criteria have been stated differently, the visibility constraint
		being replaced by a declination + Sun angle constraint. The criteria listed
		above are, however, those defining the sample.}.

@@

	\item Galactic $A_V \lesssim 0.5$ mag according to the maps of \citet{Schlegel1998}.

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
foreground bright stars. We additionally require that the Galactic extinction is
below $A_V \lesssim 0.5$ mag, based on the extinction maps by
\citet{Schlegel1998}\footnote{We use the updated values of \citet{Schlafly2011}
	to correct for foreground extinction, but the sample criterion is based on the
	old \citet{Schlegel1998} values for consistency with the first semesters of our
	program.}. Choosing low-extinction sightlines also reduce the problem of field
crowding and the contamination from Galactic outbursts posing as GRB impostors.
These additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allows
us to investigate the spectroscopic properties of GRBs.

Bursts that fulfill these selection criteria are what we define as our
\textit{statistical} sample. Using this sample we will be able to address
population properties of \swift-detected bursts. We further discuss the effect
of these selection criteria and their implication for the completeness of the
sample in Sect. \ref{completeness}.

@@

Our collaboration has set up a procedure to promptly react to alerts from
\swift~and other satellites (such as \textit{Fermi}). We process automatically
GCN notices in order to flag those events belonging to our statistical sample.
Two people are permanently on alert to manually supervise each event, for
example to recognize events of special interest beyond their inclusion in the
sample. Members of our collaboration have access to numerous facilities spread
throughout the world, among which the Nordic Optical Telescope (NOT), the Gran
Telescopio Canarias (GTC), the Telescopio Nazionale Galileo (TNG), the Gamma-Ray
Burst Optical Near-infrared Detector (GROND), the Xinlong observatory. Thanks to
this network, we can often conduct searches for optical/NIR afterglows ahead of
the time of the X-shooter spectroscopy, thus helping to plan the spectroscopic
observations. If this is not possible, we use the built-in acquisition camera of
X-shooter to search for counterparts, and adapt the observing strategy on the
fly. In many cases we could interact directly with the staff at the telescope to
aid the observation. Raw data are usually available within minutes in the ESO
archive LINK!!! and are promptly reduced by members of our collaboration, often
using archival calibration data which are readily available. This allows us in
most cases to report the preliminary results (redshift, identification of the
most prominent emission and absorption feature, etc.) within a few hours after
the beginning of the observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{RRM observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Under rare circumstances, the use of the ESO rapid response mode (RRM;
\citealt{Vreeswijk2010}) has been possible. RRM is a system which automatically
overrides current observations at the telescope. This allows the shortest
possible delay between the GRB trigger and initiation of observations. This
allows spectroscopic integration to begin on the rapidly fading, optical
transient within minutes of the burst. In case of a promptly visible GRB, a
robotic trigger was sent to the telescope if at the moment of the GCN notice the
GRB is observable and fulfills the following criteria:

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter is the currently operating instrument at the telescope.
	\item GRB promptly visible from Paranal.
	\item Error radius < 60\arcsec.
%	\item No limit on declination. 
%	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky $> 22^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation $< -12^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}. 
\end{enumerate}

The criteria to trigger RRM are looser than those to define the statistical
sample. This is because of the expected larger brightness of any GRB counterpart
minutes after the explosion, and because of the rarity of RRM triggers.

The use of RRM is unique in the sense that it allows for the detailed study of
the temporal variability of the GRB afterglow due to effect of the GRB itself on
the sorrounding medium \citep[e.g., see][]{Dessauges-Zavadsky2006,
	Vreeswijk2007, DElia2009, Vreeswijk2013}

There are nine GRBs that have been observed by us with the RRM mode. One of the
RRM triggers is outside the statistical sample  and two of the RRM trigger are
on short GRBs. The fastest response between \swift~GRB trigger and the beginning
of spectroscopic integration was for GRB~160410A, for which the delay was only
8.4 minutes. In many cases, unfortunately, we could not use RRM even for promptly visible events, because of the unavailability of X-shooter.

Potentially, the proximity of the telescope (Paranal, Chile) to the South
Atlantic Anomaly (SSA) affects the rate of RRM triggers \citep{Greiner2011}. BAT
switches off when the satellite goes through the SSA and that means we have
fewer GRB triggers immediately observable, compared to the rest of
the \swift~orbit.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which is observed with X-shooter
	and has a measured redshift, but is not a part of our statistical sample. The
	blue dots show the positions of the 61 GRBs of our statistical sample that lack
	redshift measurements. A red outline is added around the six bursts in our
	statistical sample which was observed with X-shooter, but did not yield a
	redshift measurement. Two empty circles with a red outlines indicate bursts
	outside the statistical sample that was followed up with X-shooter, but without
	a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. Please note that we removed the background where the
	sample criteria is violated ($A_V > 0.5$ mag) and replaced it with a white
	color visible at the equator. The grey-scale bar below indicates the value of
	$A_V$ on the plot. The dotted lines indicate intervals of 30$^\circ$ in
	longitude and latitude.} \label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of the two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of eight years corresponding to the ESO observing periods P84 through P98
under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 0091.C-0934, 092.A-0124,
093.A-0069, 094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055(PI:
Fynbo). These proposals were initiated on Guaranteed Time. We have included a
few additional bursts, from the programmes 084.D-0265 (PI: Benetti), 091.A-0877
(PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633 (PI: Greiner), and 095.B-0811
(PI: Levan). The total collection of spectra represents \textit{all} GRB
afterglows that have been followed up by X-shooter up to \termdate, which marks
the end of the XS-GRB legacy follow-up program.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[longauth]{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass{aa}    %% Astronomy & Astrophysics style class aa.cls v8.2
%\documentclass[referee]{aa}

@@

\usepackage{natbib}
%\usepackage{caption}

@@

\usepackage{float}
%\usepackage[symbol]{footmisc}

@@

\title{The X-shooter GRB afterglow legacy sample (XS-GRB). \thanks{Based on
		observations collected at the European Southern Observatory, Paranal, Chile,
		Program IDs: 084.A-0260, 084.D-0265, 085.A-0009, 086.A-0073, 087.A-0055,
		088.A-0051, 089.A-0067, 090.A-0088, 091.A-0877, 0091.C-0934, 092.A-0124,
		092.D-0056, 093.A-0069, 094.A-0134, 095.A-0045, 095.B-0811, 096.A-0079,
		097.A-0036, and 098.A-0055.}}

\titlerunning{The X-shooter GRB afterglow legacy sample}

@@

\abstract{In this work we present spectra of all $\gamma$-ray burst (GRB)
	afterglows that have been promptly observed with the X-shooter spectrograph
	until \termdate. In total, we obtained spectroscopic observations of 103 individual GRBs
	observed within 48 hours of the GRB trigger. Redshifts have been measured for 97
	per cent of these, covering a redshift range from 0.059 to 7.84. Based on a set
	of observational selection criteria that minimize biases with regards to
	intrinsic properties of the GRBs, the follow-up effort has been focused on
	producing a homogeneous sample of 93 afterglow spectra for GRBs discovered by
	the {\em Swift} satellite. We here provide a public release of all the reduced
	spectra, including continuum estimates and telluric absorption corrections. For
	completeness, we also provide reductions for the 18 late-time observations of
	the underlying host galaxies. We provide an assessment of the degree of
	completeness with respect to the parent GRB population, in terms of the X-ray
	properties of the bursts in the sample and find that the sample presented here
	is representative of the full \textit{Swift} sample. We constrain the fraction
	of dark bursts to be < 28 per cent and we confirm previous results that higher
	optical darkness is correlated with increased X-ray absorption. For the 42
	bursts for which it is possible, we provide a measurement of the neutral
	hydrogen column density, increasing the total number of published H{\sc I}
	column density measurements by $\sim$ 33 per cent. This dataset provides a
	unique resource to study the ISM across cosmic time, from the local progenitor
	surroundings to the intervening universe.}

\keywords{Gamma-ray burst: general --- Galaxies: high-redshift --- ISM: general --- Techniques: spectroscopic}

@@

Gamma-ray bursts (GRBs) are bright glimpses of electromagnetic radiation that
pierce through the universe, all the way from the edges of the observable
universe. They provide constraints on a very wide range of topics in
astrophysics. Examples range from small-scale phenomena relating to magnetars,
properties of highly relativistic jets, hyper/supernova explosions, the
interstellar medium, dust extinction curves, starburst galaxies, chemical and
molecular abundances, escape of ionizing radiation, the ionization state of the
intergalactic medium, intervening absorption systems to standard candles in
cosmology \citep[e.g.,][]{Wijers1998, Savaglio2006, Ghirlanda2007, Molinari2007,
	Amati2008, Vergani2009, Prochaska2009, HjorthBloom2012, Rowlinson2017,
	Christensen2017}.

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

In the beginning of the {\it Swift} era most of the follow-up afterglow
spectroscopy was secured using low-resolution spectrographs \citep[typically
$R=\lambda/\Delta\lambda$$<$1000, e.g.,][]{Fynbo2009}. Spectroscopy is powerful
as it allows us to secure information even for very faint targets
\citep{Kruhler2012}. This allows the measurement of a number of important
parameters such as redshifts, spectral slopes, and extinction. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra have been
secured, and for these events much more information about conditions inside the
host galaxies were extracted \citep[e.g.,][]{Fiore2005, Thone2007,
	Prochaska2007, Vreeswijk2007, DElia2009, Castro-Tirado2010}.

The X-shooter spectrograph \citep{Vernet2011} was the first of the second
generation instruments at the ESO Very Large Telescope (VLT). It was designed
very much with transient follow-up in mind as the fading luminosities of
such sources makes it urgent to secure as extensive wavelength coverage as possible in the
shortest possible time. At the same time, the resolution was designed to be in
the range 4000--9000 in order to be able to get a large useful spectral range
between the many sky-background emission lines in the red and near-IR spectral
ranges. The near-IR spectral coverage allows for spectroscopic observations of
the highest redshift GRBs.

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all the data resulting from the survey publicly available in reduced form (see Sect
\ref{products}).

The paper is organized in the following way: In Sect.~\ref{sample} we describe
the sample including the sample selection criteria and the observational
strategy. In Sect.~\ref{obs}, we describe the observations and the instrumental
setups, and in Sect.~\ref{proc} we detail the methodological strategies adopted
in the data reduction process and auxiliary material. In Sect.~\ref{results} we
describe the results of the survey, i.e. the efficiency of the follow-up effort
and the characteristics of the observed bursts. We also assess the
completeness of the realized sample. Finally, we offer our conclusions
in Sect.~\ref{conclusions}. We use the $\Lambda$CDM cosmology parameters
provided by the \citet{Planck2015} in which the universe is flat with $H_0 =
67.7$\,km\,s$^{-1}$ Mpc$^{-1}$ and $\Omega_\mathrm{m} = 0.307$.

@@

Being of transient nature, it is difficult to impose strong sample selection
criteria on GRBs without hampering the follow-up effort. Many natural follow-up
restrictions exist already, being it weather conditions, pointing restrictions
of the telescope or unconstrained burst localizations as reported by the alerting
facility. To maximize the return of the follow-up campaign, we have chosen a
few selection criteria that facilitate an unbiased selection of bursts, while
at the same time allowing for a high follow-up success rate. The importance of
defining unbiased selection criteria has been highlighted previously
\citep{Jakobsson2006b, Salvaterra2012, Hjorth2012, Vergani2015, Perley2016a},
when trying to address the intrinsic underlying distribution functions such as
the redshift distribution, host metallicity distribution, or afterglow
brightness distribution. When investigating a specific distribution function, a
high degree of \textit{completeness} is desired \citep[e.g.,][]{Perley2016b}.

In defining the selection criteria, we simultaneously aim to minimize any biases
against intrinsic astrophysical conditions while at the same time maximizing the
likelihood of successful observations, hence allowing us to obtain a higher
degree of completeness. By restricting the selection criteria to conditions
local to the Milky Way and therefore independent of intrinsic GRB properties,
the aim is that the collected sample represents the underlying distribution of
GRBs in a fair way. The selection criteria used here are based on previous,
similar studies \citep{Jakobsson2006b, Fynbo2009, Hjorth2012}. We characterize
the sample completeness in Sect. \ref{completeness}.

The selection criteria that define a GRB as part of our initial \textit{statistical}
	sample are:

@@

	\item GRB trigger by BAT onboard the \swift~satellite

@@

	\item The target must be visible from Cerro Paranal for at least 60 minutes, 30
degrees above the horizon, with the Sun below $-12$ degrees\footnote{Note that
	in the P84 proposal the criteria have been stated differently, the visibility
	constraint being replaced by a declination + Sun angle constraint. The criteria
	listed above are, however, those defining the sample.}.

	\item Galactic $A_V \leq 0.5$ mag according to the maps of \citet{Schlegel1998}.

@@

@@

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
foreground bright stars. We additionally require that the Galactic extinction is
below $A_V \lesssim 0.5$ mag, based on the extinction maps by
\citet{Schlegel1998}\footnote{We use the updated values of \citet{Schlafly2011}
	to correct for foreground extinction, but the sample criterion is based on the
	old \citet{Schlegel1998} values for consistency with the first semesters of our
	program.}. Choosing low-extinction sightlines also reduces the problem of field
crowding and the contamination from Galactic outbursts posing as GRB impostors.
These additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allow
us to investigate the spectroscopic properties of GRBs.

Bursts that fulfill these selection criteria are what we define as our initial
\textit{statistical} sample. To reach a higher degree of sample completeness, we
impose additional cuts that increase our ability to infer population properties.
These are presented in Sect. \ref{badbursts}. Using this sample we will be able
to address population properties of \swift-detected bursts. We further discuss
the effect of these selection criteria and their implication for the
completeness of the sample in Sect. \ref{completeness}.

@@

Our collaboration has set up a procedure to promptly react to alerts from
\swift~and other satellites (such as \textit{Fermi}). We process automatically
GCN notices in order to flag those events belonging to our statistical sample.
Two people are permanently on alert to manually supervise each event, for
example to recognize events of special interest beyond their inclusion in the
sample. Members of our collaboration have access to numerous facilities spread
throughout the world, among which the William Herschel Telescope (WHT), Nordic
Optical Telescope (NOT), the Gran Telescopio Canarias (GTC), the Telescopio
Nazionale Galileo (TNG), the Gamma-Ray Burst Optical Near-infrared Detector
(GROND \citealt{Greiner2008}), and the Xinlong observatory. Thanks to this
network, we can often conduct searches for optical/NIR afterglows ahead of the
time of the X-shooter spectroscopy, thus helping to plan the spectroscopic
observations. If this is not possible, we use the built-in acquisition camera of
X-shooter to search for counterparts, and adapt the observing strategy on the
fly. In many cases we could interact directly with the staff at the telescope to
aid the observation. Raw data are usually available within minutes in the
\href{http://archive.eso.org/wdb/wdb/eso/xshooter/form}{ESO archive} and are
promptly reduced by members of our collaboration, often using archival
calibration data which are readily available. This allows us in most cases to
report the preliminary results (redshift, identification of the most prominent
emission and absorption feature, etc.) within a few hours after the beginning of
the observations and plan additional X-shooter observations if deemed necessary.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Rapid response mode (RRM) observations} \label{RRM}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Under rare circumstances, the use of the ESO rapid response mode (RRM;
\citealt{Vreeswijk2010}) has been possible. The RRM is a system to automatically
override ongoing observations at the telescope. This allows the shortest
possible delay between the GRB trigger and the initiation of observations, where
spectroscopic integration of the rapidly fading, optical transient can commence
within minutes of the burst. One limitation is that no instrument change is
allowed in RRM, which lowers the number of successful triggers (X-shooter shares the telescope with two other instruments). In case of a promptly visible GRB, a robotic
trigger is sent to the telescope if at the time of the GCN notice the GRB
fulfills the following criteria:

\begin{enumerate}
	\item The GRB triggered \textit{Swift} onboard.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranl is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

The criteria to trigger RRM are looser than those that define the statistical
sample. This is both because of the expected larger brightness of GRB
counterparts soon after the explosion, and because of the rarity of RRM
triggers.

The use of RRM is unique as it allows to sample a long logarithmic time span in
the GRB lifetime, and it exploits the extreme brightness of early afterglows. A
pivotal exampl is the study of temporal variability of GRB afterglow absorption
systems due to effect of the GRB itself on the sorrounding medium \citep[e.g.,
see][]{Dessauges-Zavadsky2006, Vreeswijk2007, DElia2009, Vreeswijk2013}.

There are nine GRBs that have been observed with X-shooter in RRM mode. One of
the RRM triggers is outside the statistical sample, and two of the RRM triggers
are on short GRBs. Our fastest response (between the \swift~GRB trigger time and
the beginning of spectroscopic integration) was for GRB~160410A, for which the
delay was only 8.4 minutes.

In many cases, unfortunately, we could not use RRM even for promptly visible
events, e.g.{} because of the unavailability of X-shooter. Potentially, the
proximity of the telescope (Cerro Paranal, Chile) to the South Atlantic Anomaly
(SAA) affects the rate of RRM triggers \citep{Greiner2011}. BAT switches off
when the satellite goes through the SSA and that means we have fewer events
immediately observable, compared to the rest of the \swift~orbit.

@@

\begin{figure*}
	\centerline{\includegraphics{figures/skymap.pdf}} \caption{Mollweide projection
	in Galactic coordinates of the full sky showing the positions on the sky of the
	bursts presented in this work. The equator is the Galactic plane. The empty
	stars/circles are the positions of all the 1266 \textit{Swift} bursts detected
	until \termdate. Stars indicate bursts with measured redshifts and circles
	indicate those without. Blue stars indicate the position of the 104 bursts
	fulfilling the sample criteria specified in Sect. \ref{samplecrit} that have a
	measured redshift. Red outlines are added to the 84 GRBs that enter our
	statistical sample with both X-shooter spectroscopy and a measured redshift.
	Red outlines of empty stars represent bursts which have been observed with
	X-shooter and has a measured redshift, but is not a part of our statistical
	sample. The blue dots show the positions of the 61 GRBs of our statistical
	sample that lack redshift measurements. A red outline is added around the six
	bursts in our statistical sample which were observed with X-shooter, but did
	not yield a redshift measurement. Two empty circles with red outlines indicate
	bursts outside the statistical sample that were followed up with X-shooter, but
	without a redshift measurement. The different samples are compared in Sect.
	\ref{results}. The background shows the dust maps presented in
	\citet{Schlegel1998}. Please note that we removed the background where the
	sample criterion is violated ($A_V > 0.5$ mag) and replaced it with a white
	background. The grey scale bar below indicates the value of $A_V$ on the plot.
	The dotted lines indicate intervals of 30$^\circ$ in longitude and latitude.}
\label{fig:skymap}
\end{figure*}

The observations obtained for this sample have been secured with the
cross-dispersed echelle spectrograph, X-shooter \citep{Vernet2011}, mounted on
one of two Unit Telescopes at ESO/VLT, UT2 (Kueyen) and UT3 (Melipal) during
the duration of this follow-up campaign. The observations have been taken during
a period of eight years corresponding to the ESO observing periods P84 through
P98 under the following programme IDs: 084.A-0260, 085.A-0009, 086.A-0073,
087.A-0055, 088.A-0051, 089.A-0067, 090.A-0088, 092.A-0124, 093.A-0069,
094.A-0134, 095.A-0045, 096.A-0079, 097.A-0036, and 098.A-0055 (PI: Fynbo) and
0091.C-0934 (PI: Kaper). These proposals were initiated on Guaranteed Time. We
have included a few additional bursts, from the programmes 084.D-0265 (PI:
Benetti), 091.A-0877 (PI: Schady), 092.D-0056 (PI: Rau), 092.D-0633 (PI:
Greiner), and 095.B-0811 (PI: Levan). The total collection of spectra represents
\textit{all} GRB afterglows that have been followed up by X-shooter up to
\termdate, which marks the end of the XS-GRB legacy follow-up program.

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the second
Unit Telescope (UT2) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926A) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the different criteria for their selection. The
first burst observed after science verification and the mounting of X-shooter on UT2,
was GRB~091018, which is the first burst entering our statistically homogeneous
sample. For all bursts that fulfill our sample selection criteria, described in
Sect. \ref{samplecrit}, spectroscopic follow-up have been attempted with
X-shooter. Various conditions can affect our ability to follow up a given burst,
and a discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.
